---
title: "Scheduling the pipeline verbs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scheduling the pipeline verbs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`meteoTidy` does not schedule anything itself (SCOPING §9): the four pipeline
verbs -- `met_sync_live()`, `met_sync_daily()`, `met_refit()`, and
`met_backfill()` -- are ordinary R functions that you call on whatever
schedule suits your deployment. This vignette shows the recommended cadence
for each verb and a few common ways to schedule them: `cron`, GitHub Actions,
and `taskscheduleR` (Windows).

Every example below assumes you already have a site registry (a `met_site` or
`met_sites`, typically loaded via `read_sites_yaml()`) and a pipeline
configuration list naming your configured sources, e.g.:

```{r}
library(meteoTidy)

sites <- read_sites_yaml("sites.yml")

config <- list(
  store_root = "/var/meteoTidy/store",
  forecast_sources = c("openmeteo", "bom_forecast", "seasonal"),
  obs_sources = c("site_aws", "bom_obs", "silo", "ghcnh"),
  refetch_windows = list(silo = as.difftime(30, units = "days")),
  adapter_defaults = list(bom = list(allow_web_api = FALSE))
)
```

## The four verbs and their cadence

| Verb              | Cadence          | Purpose                                                     |
| ----------------- | ---------------- | ------------------------------------------------------------ |
| `met_sync_live()`  | Hourly, optional | Best-effort near-real-time obs + forecast head              |
| `met_sync_daily()` | Daily            | Full forecast refresh + curated history products            |
| `met_refit()`      | Monthly          | Recalibration (skill-gated) + partition compaction           |
| `met_backfill()`   | Ad hoc           | Day-0 bootstrap: full history, AWS export ingestion, initial fits, donor-coverage audit |

All four are multi-site (accept a single `met_site` or a `met_sites`
collection), incremental (they track their own watermarks), and idempotent
(re-running with the same inputs and clock never duplicates rows or
double-advances a watermark) -- safe to schedule directly without any
external "has this already run" bookkeeping.

```{r}
met_sync_live(sites, config = config)
met_sync_daily(sites, config = config)
met_refit(sites, config = config)
```

## The BOM hourly-vs-daily backfill trade-off (SCOPING §9)

**Every sync archives whatever forecast issuances are current when it runs**
(`archive_forecasts()`, deduplicated on `(source, model, issue_time)`), and
Open-Meteo's forecast gaps self-heal later via its Previous/Single-Runs
endpoints -- but **BOM issues forecasts several times a day, and a missed
issuance cannot be backfilled** (there is no historical-issuance API for
BOM's products). This means:

- **Hourly `met_sync_live()` users** capture most of BOM's intra-day issue
  cycles, because the live sync polls (and archives) roughly as often as BOM
  re-issues.
- **Daily-only users** (running only `met_sync_daily()`) get one forecast
  snapshot per day and **permanently miss** any BOM issuance that occurred
  and was superseded between two daily runs -- there is no way to recover it
  after the fact.

If your deployment cares about a complete BOM issuance history, schedule
`met_sync_live()` hourly (or as close to it as your BOM transport budget
allows) in addition to the daily job. If you only need daily snapshots (e.g.
you mainly consume `history_daily`/`history_hourly` and Open-Meteo
forecasts), a daily-only schedule is a reasonable, explicit trade-off --
just don't expect a complete BOM archive from it.

## `cron` (Linux/macOS)

A typical crontab running an `Rscript` wrapper per verb:

```sh
# m h dom mon dow command
0    * * * *  Rscript /opt/meteoTidy/run-sync-live.R    >> /var/log/meteoTidy/live.log    2>&1
15   2 * * *  Rscript /opt/meteoTidy/run-sync-daily.R   >> /var/log/meteoTidy/daily.log   2>&1
30   3 1 * *  Rscript /opt/meteoTidy/run-refit.R         >> /var/log/meteoTidy/refit.log   2>&1
```

Each `run-*.R` script loads the site registry and config, then calls exactly
one verb -- e.g. `run-sync-live.R`:

```{r}
#!/usr/bin/env Rscript
sites  <- meteoTidy::read_sites_yaml("/etc/meteoTidy/sites.yml")
config <- yaml::read_yaml("/etc/meteoTidy/pipeline-config.yml")
meteoTidy::met_sync_live(sites, config = config)
```

## GitHub Actions

A scheduled workflow using `r-lib/actions` to set up R, running the daily
sync and committing/pushing an updated store (or, more commonly, pushing to
external storage such as S3/GCS instead of the repo itself):

```yaml
name: met-sync-daily
on:
  schedule:
    - cron: "15 2 * * *"
  workflow_dispatch: {}

jobs:
  sync:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: r-lib/actions/setup-r@v2
      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: any::meteoTidy
      - name: Run daily sync
        run: Rscript -e 'meteoTidy::met_sync_daily(meteoTidy::read_sites_yaml("sites.yml"), config = yaml::read_yaml("pipeline-config.yml"))'
```

A separate hourly workflow (`cron: "0 * * * *"`) runs `met_sync_live()` the
same way; a separate monthly workflow (`cron: "30 3 1 * *"`) runs
`met_refit()`.

## `taskscheduleR` (Windows)

For a Windows host, `taskscheduleR::taskscheduler_create()` registers a
Windows Task Scheduler job that runs an R script on a recurring schedule:

```{r}
library(taskscheduleR)

taskscheduler_create(
  taskname = "meteoTidy-sync-live",
  rscript = "C:/meteoTidy/run-sync-live.R",
  schedule = "HOURLY"
)

taskscheduler_create(
  taskname = "meteoTidy-sync-daily",
  rscript = "C:/meteoTidy/run-sync-daily.R",
  schedule = "DAILY",
  starttime = "02:15"
)

taskscheduler_create(
  taskname = "meteoTidy-refit",
  rscript = "C:/meteoTidy/run-refit.R",
  schedule = "MONTHLY",
  startdate = format(Sys.Date(), "%d/%m/%Y")
)
```

## `met_backfill()` is not scheduled

`met_backfill()` is an ad hoc, day-0 bootstrap: run it once when onboarding a
new site (or re-run it manually after a long outage), not on a recurring
schedule. It pulls full history, ingests any historical AWS logger exports
you supply, makes initial calibration fits, and returns a donor-coverage
audit you should read before relying on the gap-fill donor ladder for that
site (SCOPING §13).

```{r}
met_backfill(
  new_site,
  config = config,
  aws_export = "exports/new_site-logger.csv"
)
```

## Production setup: hourly live + daily sync

This is the configuration used for the Katoomba (`kat`) and Blaxland
(`blax`) waste-management facilities. It is the one the live acceptance
script (`inst/acceptance/live_sync.R`) checks.

### Site YAML

The full file ships as `system.file("acceptance", "sites.yaml", package =
"meteoTidy")`; replace `@STORE_ROOT@` with the store directory. It holds no
secrets: API keys are named by environment variable only. One site:

```yaml
sites:
  - site_id: kat
    latitude: -33.70
    longitude: 150.32
    elevation: 1000
    timezone: Australia/Sydney
    store_root: D:/meteo/store
    resolved:
      bom:
        geohash: r64bhq        # BOM web-API location (forecast + obs)
        aac: NSW_PT072         # Katoomba, précis product IDN11060
        wmo: "94743"           # Mount Boyce AWS, obs product IDN60901
    sources:
      eagleio:                 # listed first: it wins over bom_obs in history
        adapter: eagleio
        api_key_env: EAGLE_API_KEY
        nodes:
          precipitation: 64642e52fbaed638fdb04110
          temperature_2m: 64642e52fbaed638fdb04100
      bom_obs:     {adapter: bom_obs, allow_web_api: true}
      silo:        {adapter: silo, dataset: data_drill, api_key_env: SILO_API_KEY}
      openmeteo:   {adapter: openmeteo, product: forecast}   # ECMWF IFS, GFS, ICON
      om_ens:      {adapter: openmeteo, product: ensemble}   # ECMWF IFS + ICON ensembles
      bom_forecast: {adapter: bom_forecast, allow_web_api: true}
```

- `openmeteo` archives three named global models -- `ecmwf_ifs025`,
  `gfs_global`, `icon_global` -- each stamped with its real run time from
  Open-Meteo's metadata. `best_match` is not archived unless listed in
  `models:`, and then it is flagged as having no verifiable run time.
- `om_ens` archives the ECMWF IFS (51 members, 15 days) and ICON (40
  members, 7.5 days) ensembles.
- A run already archived in full is not downloaded again, so hourly polling
  costs one small metadata request per model between runs.
- `bom_forecast` takes the daily forecast from BOM's web API for the site's
  own geohash; the town précis (`aac:`) is only a fallback, and its rows are
  labelled `model = "daily_precis"`.
- The free-tier notice is shown once per R session. For a commercial
  Open-Meteo plan add `api_key_env: OPEN_METEO_KEY` to both Open-Meteo
  sources; every request then goes to the `customer-` host with the key.

Blaxland is the same, with `latitude: -33.73`, `longitude: 150.61`,
`elevation: 250`, `geohash: r65050`, `aac: NSW_PT129` (Springwood),
`wmo: "94763"` (Penrith Lakes), and eagle.io nodes
`precipitation: 647e9492dac9bf46aa1ffd0f` and
`temperature_2m: 647e9492dac9bf46aa1ffd02`.

### Config

```{r}
store <- "D:/meteo/store"

live_config <- list(
  store_root = store,
  obs_sources = c("bom_obs", "eagleio"),
  forecast_sources = c("openmeteo", "om_ens", "bom_forecast"),
  lock_timeout = 1800            # seconds to wait for an overlapping run
)

daily_config <- list(
  store_root = store,
  obs_sources = c("bom_obs", "eagleio", "silo"),
  forecast_sources = c("openmeteo", "om_ens", "bom_forecast"),
  refetch_windows = list(
    silo = as.difftime(30, units = "days"),    # SILO revises recent days
    eagleio = as.difftime(7, units = "days"),
    bom_obs = as.difftime(3, units = "days")   # the 72-hour product
  ),
  lock_timeout = 1800
)
```

Set `SILO_API_KEY` (the SILO email address) and `EAGLE_API_KEY` in the
scheduler's environment.

### Schedule

- **`met_sync_live()` hourly, at 20 minutes past the hour.** BOM's
  half-hourly observations and hourly forecast re-issues are in by then.
  Open-Meteo runs every 6 hours and the archive keeps one copy per run, so
  hourly polling costs little.
- **`met_sync_daily()` once a day at 10:30 Australia/Sydney.** SILO's
  previous day closes at 9 am local, and the 30-day refetch window picks up
  later SILO revisions.
- **`met_compact()` weekly, e.g. Sunday 03:40.** Every sync appends small
  Parquet files (one per partition written, plus one per site to the
  `qc_log` and `obs_transport` logs). Compaction rewrites each partition as
  one file; reads return the same rows before and after.

All three lock `store_root`. If a slow hourly run overlaps the next one (or
the compaction), the second waits for the lock (up to `lock_timeout`)
rather than writing duplicates, so there is no need for the scheduler to
prevent overlap.

Each HTTP request times out after 60 s (`options(meteoTidy.http_timeout =)`),
is retried with backoff, and then fails only its own source, so a hung
upstream cannot hold the lock indefinitely.

The production `Rscript` entry points:

```{r}
# run-sync-live.R -- hourly at hh:20
library(meteoTidy)
sites <- read_sites_yaml("D:/meteo/sites.yaml")
met_sync_live(sites, config = live_config, fail_on = "failed")
```

```{r}
# run-sync-daily.R -- daily at 10:30 Australia/Sydney
library(meteoTidy)
sites <- read_sites_yaml("D:/meteo/sites.yaml")
met_sync_daily(sites, config = daily_config, fail_on = "failed")
```

```{r}
# run-compact.R -- weekly, e.g. Sunday 03:40
library(meteoTidy)
print(met_compact("D:/meteo/store", lock_timeout = 1800))
```

With `taskscheduleR`:

```{r}
taskscheduleR::taskscheduler_create(
  taskname = "meteoTidy-compact", rscript = "D:/meteo/run-compact.R",
  schedule = "WEEKLY", days = "SUN", starttime = "03:40"
)
```

### Detecting failure

Each run writes one line per site to **stderr**, for example:

```
meteoTidy met_sync_daily kat: ok | bom_obs ok 1332 | eagleio ok 2016 | silo ok 150 | openmeteo ok 8998 | om_ens ok 107304 | bom_forecast ok 908
meteoTidy met_sync_live blax: degraded | bom_obs ok 884 | eagleio STALE (eagle.io station is stale: no data from "eagleio" for the requested window. ℹ Last value reported 2026-09-24 03:00 UTC.) | openmeteo ok 8998 | om_ens ok 107304 | bom_forecast ok 908
```

`fail_on` decides when the process exits non-zero:

| `fail_on` | Exit status is non-zero when |
| --------- | ---------------------------- |
| `"none"` (default) | never; inspect the returned table instead |
| `"failed"` (recommended) | any source `"failed"` or any site `"error"`; a `"stale"` source alone does not |
| `"all"`   | every site failed outright (nothing was acquired) |
| `"any"`   | any site is not fully `"ok"`, including one stale or failed source |

Use `"failed"` in production, and log stderr for the per-source detail. A
station that is offline (as Blaxland's eagle.io logger has been since
24 September 2026) is `"stale"`, which is not a fault in meteoTidy or the
feed, so `"failed"` stays quiet about it, while `"any"` would alarm on every
run until it is fixed. `"all"` misses a dead feed as long as anything else
works.

The returned tibble has the same information, structured: `status` per site
and a `sources` list-column with each source's `status` (`"ok"`,
`"failed"`, `"stale"`), row count `n` and `message`. With `fail_on`, the
error condition (`meteoTidy_error_sync_failed`) carries that table as
`cnd$status`.

### Reading for meteoHazard

`met_wide()` serves one source per call, and each variable from one model,
never a blend. `model` is a precedence list: each variable comes from the
first listed model that has it (by default ECMWF IFS, then GFS, then ICON,
so boundary-layer height comes from GFS and soil moisture from ICON).
`met_provenance()` names the model and statistic behind each column.

```{r}
next72 <- list(from = now, to = now + 71 * 3600)
met <- met_wide(sites@sites[[1]], next72)                           # Open-Meteo, per-variable models
bom <- met_wide(sites@sites[[1]], next72, source = "bom_forecast", model = "hourly")
gust95 <- met_wide(sites@sites[[1]], next72, source = "om_ens",      # 95th percentile of members
                   variables = c("wind_gusts_10m", "temperature_2m"), stat = "p95")
rain90 <- met_wide(sites@sites[[1]], next72, source = "bom_forecast", model = "hourly",
                   variables = "precipitation", stat = "p90")      # BOM's 10 % chance amount
```

For the email's other fields, pass `variables` explicitly, e.g.
`c("weather_code", "is_day", "uv_index", "cape", "wind_speed_80m")`;
text fields (précis, `fire_danger_category`, UV category) are in
`met_forecast_aux()`.
