---
title: "Scheduling the pipeline verbs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scheduling the pipeline verbs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`meteoTidy` does not schedule anything itself (SCOPING §9): the four pipeline
verbs -- `met_sync_live()`, `met_sync_daily()`, `met_refit()`, and
`met_backfill()` -- are ordinary R functions that you call on whatever
schedule suits your deployment. This vignette shows the recommended cadence
for each verb and a few common ways to schedule them: `cron`, GitHub Actions,
and `taskscheduleR` (Windows).

Every example below assumes you already have a site registry (a `met_site` or
`met_sites`, typically loaded via `read_sites_yaml()`) and a pipeline
configuration list naming your configured sources, e.g.:

```{r}
library(meteoTidy)

sites <- read_sites_yaml("sites.yml")

config <- list(
  store_root = "/var/meteoTidy/store",
  forecast_sources = c("openmeteo", "bom_forecast", "seasonal"),
  obs_sources = c("site_aws", "bom_obs", "silo", "ghcnh"),
  refetch_windows = list(silo = as.difftime(30, units = "days")),
  adapter_defaults = list(bom = list(allow_web_api = FALSE))
)
```

## The four verbs and their cadence

| Verb              | Cadence          | Purpose                                                     |
| ----------------- | ---------------- | ------------------------------------------------------------ |
| `met_sync_live()`  | Hourly, optional | Best-effort near-real-time obs + forecast head              |
| `met_sync_daily()` | Daily            | Full forecast refresh + curated history products            |
| `met_refit()`      | Monthly          | Recalibration (skill-gated) + partition compaction           |
| `met_backfill()`   | Ad hoc           | Day-0 bootstrap: full history, AWS export ingestion, initial fits, donor-coverage audit |

All four are multi-site (accept a single `met_site` or a `met_sites`
collection), incremental (they track their own watermarks), and idempotent
(re-running with the same inputs and clock never duplicates rows or
double-advances a watermark) -- safe to schedule directly without any
external "has this already run" bookkeeping.

```{r}
met_sync_live(sites, config = config)
met_sync_daily(sites, config = config)
met_refit(sites, config = config)
```

## The BOM hourly-vs-daily backfill trade-off (SCOPING §9)

**Every sync archives whatever forecast issuances are current when it runs**
(`archive_forecasts()`, deduplicated on `(source, model, issue_time)`), and
Open-Meteo's forecast gaps self-heal later via its Previous/Single-Runs
endpoints -- but **BOM issues forecasts several times a day, and a missed
issuance cannot be backfilled** (there is no historical-issuance API for
BOM's products). This means:

- **Hourly `met_sync_live()` users** capture most of BOM's intra-day issue
  cycles, because the live sync polls (and archives) roughly as often as BOM
  re-issues.
- **Daily-only users** (running only `met_sync_daily()`) get one forecast
  snapshot per day and **permanently miss** any BOM issuance that occurred
  and was superseded between two daily runs -- there is no way to recover it
  after the fact.

If your deployment cares about a complete BOM issuance history, schedule
`met_sync_live()` hourly (or as close to it as your BOM transport budget
allows) in addition to the daily job. If you only need daily snapshots (e.g.
you mainly consume `history_daily`/`history_hourly` and Open-Meteo
forecasts), a daily-only schedule is a reasonable, explicit trade-off --
just don't expect a complete BOM archive from it.

## `cron` (Linux/macOS)

A typical crontab running an `Rscript` wrapper per verb:

```sh
# m h dom mon dow command
0    * * * *  Rscript /opt/meteoTidy/run-sync-live.R    >> /var/log/meteoTidy/live.log    2>&1
15   2 * * *  Rscript /opt/meteoTidy/run-sync-daily.R   >> /var/log/meteoTidy/daily.log   2>&1
30   3 1 * *  Rscript /opt/meteoTidy/run-refit.R         >> /var/log/meteoTidy/refit.log   2>&1
```

Each `run-*.R` script loads the site registry and config, then calls exactly
one verb -- e.g. `run-sync-live.R`:

```{r}
#!/usr/bin/env Rscript
sites  <- meteoTidy::read_sites_yaml("/etc/meteoTidy/sites.yml")
config <- yaml::read_yaml("/etc/meteoTidy/pipeline-config.yml")
meteoTidy::met_sync_live(sites, config = config)
```

## GitHub Actions

A scheduled workflow using `r-lib/actions` to set up R, running the daily
sync and committing/pushing an updated store (or, more commonly, pushing to
external storage such as S3/GCS instead of the repo itself):

```yaml
name: met-sync-daily
on:
  schedule:
    - cron: "15 2 * * *"
  workflow_dispatch: {}

jobs:
  sync:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4
      - uses: r-lib/actions/setup-r@v2
      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: any::meteoTidy
      - name: Run daily sync
        run: Rscript -e 'meteoTidy::met_sync_daily(meteoTidy::read_sites_yaml("sites.yml"), config = yaml::read_yaml("pipeline-config.yml"))'
```

A separate hourly workflow (`cron: "0 * * * *"`) runs `met_sync_live()` the
same way; a separate monthly workflow (`cron: "30 3 1 * *"`) runs
`met_refit()`.

## `taskscheduleR` (Windows)

For a Windows host, `taskscheduleR::taskscheduler_create()` registers a
Windows Task Scheduler job that runs an R script on a recurring schedule:

```{r}
library(taskscheduleR)

taskscheduler_create(
  taskname = "meteoTidy-sync-live",
  rscript = "C:/meteoTidy/run-sync-live.R",
  schedule = "HOURLY"
)

taskscheduler_create(
  taskname = "meteoTidy-sync-daily",
  rscript = "C:/meteoTidy/run-sync-daily.R",
  schedule = "DAILY",
  starttime = "02:15"
)

taskscheduler_create(
  taskname = "meteoTidy-refit",
  rscript = "C:/meteoTidy/run-refit.R",
  schedule = "MONTHLY",
  startdate = format(Sys.Date(), "%d/%m/%Y")
)
```

## `met_backfill()` is not scheduled

`met_backfill()` is an ad hoc, day-0 bootstrap: run it once when onboarding a
new site (or re-run it manually after a long outage), not on a recurring
schedule. It pulls full history, ingests any historical AWS logger exports
you supply, makes initial calibration fits, and returns a donor-coverage
audit you should read before relying on the gap-fill donor ladder for that
site (SCOPING §13).

```{r}
met_backfill(
  new_site,
  config = config,
  aws_export = "exports/new_site-logger.csv"
)
```
